Efficient static analysis of Marlowe contracts

### 2020-05-28 papers/static_analysis2020/paper.tex
Various ods: revised conclusions, put future work into conclusions. Space saving in Section 5 and Section 6.

Thanks to static analysis, we can automatically check beforehand whether any
payments promised by a Marlowe~\cite{Marlowe:2018,marlowe} contract can be fulfilled in every
possible execution of the contract. If a Marlowe contract has passed
the static analysis, we will have a very high assurance
that whenever the contract says it will make a payment, the contract will
indeed have enough money available.

Marlowe's static analysis relies on SMT solvers, which are able to check efficiently
whether a set of constraints is satisfiable. Thanks to state of the art
libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
high level language. In the case of SBV, we can write properties as
Haskell functions, with few restrictions on how those functions are
implemented; SBV automatically translates those functions to SMTLib format \cite{barrett2010smt}, a language many SMT solvers understand.

Nevertheless, static analysis can become very computationally expensive if
implemented naïvely, because it is an NP-complete problem in the general case.
In our case the main property is whether a contract will have enough money for all payments. Indeed, high level abstractions are often a tradeoff with efficiency.

@@

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches, an approach to ensure correctness
of the optimisations that combines the usage of property-based testing
and verification using automated theorem provers, and empirical data
that measures the effect of some of the optimisations on the implementation
of our case-study: the optimisation of Marlowe's static analysis.

The techniques described helped us reduce the analysis time of an implementation
that followed the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract takes less
than a second to analyse, and where a four-person crowdfunding contract that
fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes.

We have classified optimisation techniques as lightweight and heavyweight.

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. But these optimisations can also
translate in important reductions to execution time and memory usage, as
shown by the experiments we report in Section~\ref{sec:Measurements}.
We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we only
represent each equivalence class of executions once, i.e:
we represent them in a normal form.
\item \textbf{Minimizing the representation of inputs and outputs}. We encode inputs
and outputs as concisely as possible, discarding inferable information.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we illustrate the use of property based testing on the implementation with
heavyweight optimisations. Finally, we present empirical results that
show the effect of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis (Section~\ref{sec:Measurements}).

@@

To sum up, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters}, we already limitted the number of inputs per transaction
to a maximum of one. But now we have also assigned each of the
transactions to a part of the contract, so we no longer need a symbolic
list of transactions, we can use a finite concrete list of symbolic
transactions.

One detail we found out during testing is that, even though a \texttt{When}
belongs to the begining of a transaction, the environment used by
the \texttt{Observation}s in the \texttt{Notify} cases of the \texttt{When}
correspond to the \texttt{State} before the \texttt{When} is executed (except
for the slot interval). The same is true for the \texttt{Value} in the
\texttt{Deposit} cases, since the amount to deposit must be calculated without
considering the effects of the deposit itself.

@@

We use this separation of concerns between the static analysis and
the concrete semantics to our advantage, as an opportunity for applying
property based testing, as we explain in Section~\ref{subsec:Testing-for-consistency}.

@@

Errors often occur when optimising or translating. If the static analysis
implementation is close to the actual implementation, it is much less
likely that a mistake will be made when writing it. On the other hand,
if we have to craft a static analysis implementation that is equivalent
(in some sense) to the original, but is written in a very different
way, it is more likely that we will make assumptions which do not
hold.

In order to avoid these problems, we combine the use of automated proof
assistants to verify our assumptions and the use of property based testing to
test for consistency and equivalence (as we describe in
Section~\ref{subsec:Testing-for-consistency}).

@@

A limitation of this approach is that it only tests for false positives,
and it does not prevent false negatives, but those can be covered by testing
for equivalence (see Section~\ref{subsec:Testing-for-equivalence}).

In addition, we can add assertions to the process. In the case of
Marlowe, if the counterexample causes errors during the execution or is
formed incorrectly, it would also mean that there is a problem with
the static analysis. For example, it may be that the counterexample
refers to a \texttt{Case} of a \texttt{When} that does not exist,
or that it has invalid or ambiguous intervals. If it has \texttt{UselessTransactions}
is ok, because we are doing that on purpose, as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
    \begin{centering}
    \includegraphics[width=0.8\textwidth]{figures/consistency_property}
    \par\end{centering}
    \caption{Testing for consistency\label{fig:Property-based-testing-for-consistency}}
\end{minipage}%
\begin{minipage}{.5\textwidth}
    \begin{centering}
    \includegraphics[width=0.7\textwidth]{figures/equivalence_property}
    \par\end{centering}
    \caption{Testing for equivalence\label{fig:Property-based-testing-for-equivalence}}
\end{minipage}
\end{figure}

@@

If we have two static analysis implementations, we have another opportunity
for testing. In our case, we have one efficient implementation that
is very different from the semantics and one inefficient implementation
that is much closer to the semantics, and we can compare the results of the
two implementations. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
We generate random contracts using a custom QuickCheck \cite{quickcheck} generator,
we feed them to both static analysis implementations, and we compare the results.
If the results are the same then test passes, if they are different then
one of the implementations is wrong.

This approach covers both types of errors, i.e: false positives and
false negatives (for both of the implementations), but the execution
time of the tests is bounded from below by the slower of the two implementations.
The consistency approach is thus more efficient in finding false positives.

@@

Using the \texttt{perf} tool \cite{perf}, we have measured the execution
time\footnote{The experiments were run on a laptop computer with a
i9-9900K (3.6GHz) processor and two modules of 16GB of SODIMM DDR4 RAM at 2400MHz.}
and the overhead of both the generation of the constraints and their solution
by Z3 \cite{de2008z3}. And we have measured the peak RAM usage of the whole
process by using GNU's \texttt{time} tool \cite{keppel1998gnu}.

In all the contracts, the implementation that has the heavyweight optimisations
performs much better and scales further. In the case of the \textit{auction}
and \textit{crowdfunding} contracts, whose size grows exponentially, both
approaches quickly overwhelm the resources available in the execution environment.
On the other hand, in the case of \textit{rent} and \textit{coupon bond} contracts,
which grow linearly, by using the lightweight version, the problem becomes 
intractable much faster than with the heavyweight version.

@@

These results suggest that the SBV library is able to generate constraints in a way
that they are handled efficiently by Z3, but the process itself can be costly
if we are not careful. Nevertheless, the overall execution time required by Z3
is also lower in the case of the version with heavyweight optimisations and it
grows more slowly, which suggests the optimisations described in this paper
affect both stages of the process to some extent.

@@

Work in \cite{mokhov2019formal} documents a similar effort to ensure
correctness of implementations by using Haskell and SBV library; the
authors also discuss performance of the analysis and apply this approach
to non-functional requirements.

The idea of using constraint solvers for finding bugs is not new,
and there has been a number of initiatives that have explored its
application to the verification of assertions in programs written
using general purpose programming languages \cite{gulwani2008program,jackson2000finding};
as well as for the compliance with usage protocols
\cite{ball2001automatically,xie2005saturn}.

@@

\section{Future work}

In the future, we would like to extend static analysis to cover other
potential problems in Marlowe contracts and to aid their development.
We can envisage that we can use static analysis to locate unreachable
subcontracts, to allow developers to provide custom assertions and
check their satisfiability, and even to allow users to inspect the
possible maximum and minimum values that particular expressions can
reach.

@@

In this paper, we have summarized the lessons learned from our work
on optimising the static analysis for Marlowe contracts. We have found
that there are two clear approaches to static analysis using SMT,
both have advantages and disadvantages. One is less error prone and
straightforward, but inefficient and hard to test; the other approach
is much more efficient, versatile, and testable, but more error prone.

We have also seen that many specific properties and restrictions characteristic
of the target DSL can be utilized both as optimisation opportunities
and, in our case, for completeness of the analysis. Symbolic execution
of, for example, a turing complete language, would be intractable,
and would require us to manually set a bound; but this is not in the
case for Marlowe.

In the end, we have illustrated how to counteract the main disadvantage
of the later approach, its propensity to errors, by using property
based testing. This way we have obtained a static analysis implementation
that is efficient, versatile, testable, and reliable.

On the other hand, for the static analysis of Marlowe contracts, we
found out that when running statistics on the equivalence testing
property, most of the bugs were false negatives in the straightforward
implementation, and the optimised implementation seems to be more
reliable thanks to the consistency tests that we run beforehand.

Another advantage of the optimised implementation is that, because
it relies on fewer and simpler features, it is compatible with more
SMT solvers which, in turn, means that it is less reliant on the
correctness of a single SMT solver. If an SMT solver fails to give
an answer, we can try a different one; if we want to ensure that a
contract is valid, we can test it with several SMT solvers and ensure
all of them agree.

>>>>

Thanks to static analysis, we can automatically check beforehand whether any
payments promised by a Marlowe~\cite{marlowe2020,Marlowe:2018} contract can be fulfilled in every
possible execution of the contract. If a Marlowe contract has passed
the static analysis, we will have a very high assurance
that whenever the contract says it will make a payment, the contract will
indeed have enough money available.

Marlowe's static analysis relies on SMT solvers, which are able to check efficiently
whether a set of constraints is satisfiable. The main property for us is whether a contract will have enough money for all payments to be made in full.

Thanks to state of the art
libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
high level language. In the case of SBV, we can write properties as
Haskell functions, with few restrictions on how those functions are
implemented; SBV automatically translates those functions to SMTLib format \cite{barrett2010smt}, a language many SMT solvers understand.
However, high level abstractions often have a tradeoff with efficiency, and static analysis can become very expensive computationally if
implemented naïvely, because it is NP-complete in the general case.

% Indeed, high level abstractions are often a tradeoff with efficiency.

@@

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches. Our approach aims to ensure correctness
of the optimisations by combining the use of property-based testing
and with verification in an automated theorem prover to establish properties of the optimisations. We also present empirical data
that measures the effect of some of the optimisations on the implementation
of our Marlowe case-study.

The techniques described helped us reduce the analysis time of an implementation
that followed the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract takes less
than a second to analyse, and where a four-person crowdfunding contract that
fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes. 
We have classified optimisation techniques as lightweight and heavyweight.

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. But these optimisations can also
translate in important reductions to execution time and memory usage, as
shown by the experiments we report in Section~\ref{sec:Measurements}.
We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we only
represent each equivalence class of executions once, i.e:
we represent them in a normal form.
\item \textbf{Minimizing the representation of inputs and outputs}. We encode inputs
and outputs as concisely as possible, discarding inferable information.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), and we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}).
We then, in Section \ref{sec:Making-static-analysis-efficient} explore in Section the lightweight and heavyweight 
optimisation techniques in more detail, and illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we illustrate the use of property based testing on the implementation with
heavyweight optimisations. Finally, we present empirical results that
show the effect of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis (Section~\ref{sec:Measurements}).

@@

To sum up, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters}, we already limited the number of inputs per transaction
to a maximum of one. But now we have also assigned each of the
transactions to a part of the contract, so we no longer need a symbolic
list of transactions, we can use a finite concrete list of symbolic
transactions.

One detail we found out during testing is that, even though a \texttt{When}
belongs to the beginning of a transaction, the environment used by
the \texttt{Observation}s in the \texttt{Notify} cases of the \texttt{When}
correspond to the \texttt{State} before the \texttt{When} is executed (except
for the slot interval). The same is true for the \texttt{Value} in the
\texttt{Deposit} cases, since the amount to deposit must be calculated without
considering the effects of the deposit itself.

@@

We use this separation of concerns between static analysis and
concrete semantics to our advantage, as an opportunity for applying
property based testing.%, as we explain in Section~\ref{subsec:Testing-for-consistency}.

@@

Errors often occur when optimising or translating. If the static analysis
implementation is close to the actual implementation, it is much less
likely that a mistake will be made when writing it. On the other hand,
if we have to craft a static analysis implementation that is equivalent
(in some sense) to the original, but is written in a very different
way, it is more likely that we will make assumptions which do not
hold.
In order to avoid these problems, we combine the use of automated proof
assistants to verify our assumptions and the use of property based testing to
test for consistency and equivalence. %as we describe in
%Section~\ref{subsec:Testing-for-consistency}).

@@

A limitation of this approach is that it only tests for false positives; 
false negatives can be detected by testing
for equivalence (see Section~\ref{subsec:Testing-for-equivalence}).

In addition, we can add assertions to the process. In the case of
Marlowe, if the counterexample causes errors during the execution or is
formed incorrectly, it would also mean that there is a problem with
the static analysis. For example, it may be that the counterexample
refers to a \texttt{Case} of a \texttt{When} that does not exist,
or that it has invalid or ambiguous intervals. If it has \texttt{UselessTransactions}
that is ok, because we are doing that on purpose, as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.

\begin{figure}[h]
\centering
\begin{minipage}{.4\textwidth}
    \begin{centering}
    \includegraphics[width=0.8\textwidth]{figures/consistency_property}
    \par\end{centering}
    \caption{Testing for consistency\label{fig:Property-based-testing-for-consistency}}
\end{minipage}%
\begin{minipage}{.4\textwidth}
    \begin{centering}
    \includegraphics[width=0.7\textwidth]{figures/equivalence_property}
    \par\end{centering}
    \caption{Testing for equivalence\label{fig:Property-based-testing-for-equivalence}}
\end{minipage}
\end{figure}

\vspace*{-5mm}

@@

Given two  implementations of the static analysis, we have another opportunity
for testing. We have one efficient implementation that
is very different from the semantics and one inefficient implementation
that is much closer to the semantics, and we can compare the results of the
two, as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.

We generate random contracts using a custom QuickCheck \cite{quickcheck} generator,
 feed them to both implementations, and compare the results.
If the results are the same then test passes, if they are different then
one of the implementations is wrong.
This approach covers both types of errors, i.e: false positives and
false negatives (for both of the implementations), but the execution
time of the tests is bounded from below by the slower of the two implementations.
The consistency approach is thus more efficient in finding false positives.

@@

Using the \texttt{perf} tool \cite{perf}, we measured the execution
time\footnote{The experiments were run on a laptop computer with a
i9-9900K (3.6GHz) processor and two modules of 16GB of SODIMM DDR4 RAM at 2400MHz.}
and the overhead of the generation of the constraints and their solution
by Z3 \cite{de2008z3}. We also measured the peak RAM usage of the whole
process using GNU's \texttt{time} tool \cite{keppel1998gnu}.

In all cases the implementation with the heavyweight optimisations
performs much better and scales further. In the case of the \textit{auction}
and \textit{crowdfunding} contracts, whose size grows exponentially, both
approaches quickly overwhelm the resources available in the execution environment.
In the case of \textit{rent} and \textit{coupon bond} contracts,
which grow linearly, when using the lightweight version the problem becomes 
intractable much faster than with the heavyweight one.

@@

These results suggest that the SBV library is able to generate constraints in a way
that they are handled efficiently by Z3, but the process itself can be costly. However, the  execution time required by Z3
is also lower in the case of the heavyweight optimisations as well as
growing more slowly, which suggests that the optimisations described here
affect both parts of the process.

@@

Work in \cite{mokhov2019formal} documents a similar effort to ensure
correctness of control software in Haskell using the SBV library; the
authors also discuss performance of the analysis and apply this approach
to non-functional requirements.

The idea of using constraint solvers for finding bugs is not new,
and there have been a number of initiatives that have explored its
application to the verification of assertions in programs written
using general purpose programming languages \cite{gulwani2008program,jackson2000finding};
as well as for the compliance with protocols
\cite{ball2001automatically,xie2005saturn}.

@@

@@

In this paper we have summarized our work
on optimising the static analysis for Marlowe contracts. We have seen
that there are two distinct approaches to static analysis using SMT,
both have advantages and disadvantages. One is less error prone and
straightforward, but inefficient and hard to test; the other 
is much more efficient, versatile, and testable, but more error prone.
We have also seen that many specific properties and restrictions characteristic
of the target DSL can be utilized both as optimisation opportunities
and, in our case, for completeness of the analysis. Symbolic execution
of a Turing-complete language, would be intractable,
and would require us to manually set a bound; but this is not in the
case for Marlowe.

In the end, we have illustrated how to counteract the main disadvantage
of the optimised approach -- its propensity to errors -- by using property
based testing. This way we have obtained a static analysis implementation
that is efficient, versatile, testable, and reliable. 
On the other hand, for the static analysis of Marlowe contracts, we
found out that when running statistics on the equivalence testing
property, most of the bugs were false negatives in the straightforward
implementation, and the optimised implementation seems to be more
reliable thanks to the consistency tests that we run beforehand.

Another advantage of the optimised implementation is that, because
it relies on fewer and simpler features, it is compatible with more
SMT solvers which, in turn, means that it is less reliant on the
correctness or efficiency of a single solver. If one solver fails to give
an answer, we can try another; if we want further evidence that a
contract is valid, we can test it with several solvers.

In the future, we would like to extend static analysis to cover other
potential problems in Marlowe contracts and to aid their development.
We plan to use static analysis to locate unreachable
subcontracts, to allow developers to provide custom assertions and
check their satisfiability, and to allow users to inspect the
possible maximum and minimum values that particular expressions can
reach.
